---
title: "Computational_Musicology_dashboard"
author: "Gijs Schouten"
date: "2025-02-12"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Dashboard



### Plot 1

``` {r}

library(ggplot2)
library(dplyr)


corpus <- read.csv("aisc.csv", header = TRUE)


corpus_avg <- corpus %>%
  group_by(arousal) %>%
  summarise(avg_danceability = mean(danceability, na.rm = TRUE))

ggplot(corpus_avg, aes(x = arousal, y = avg_danceability)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  theme_minimal() +
  labs(
    title = "Arousal vs Danceability (Trend Line)",
    x = "Arousal (Energy Level)",
    y = "Average Danceability"
  )


```

***

X-axis (Arousal - Energy Level): Measures the intensity/energy of the song. Higher values mean the song feels more energetic.
Y-axis (Danceability): Measures how easy the song is to dance to.

This plot visualizes the relationship between arousal (energy level) and danceability. The fluctuating trend suggests that there is no direct correlation. 
Some high-energy songs are highly danceable, while others are not. Peaks around arousal levels 4.0 and 4.5 indicate that certain energy levels favor danceability, but the sudden drops highlight that not all energetic songs are easy to dance to.

### Plot 2

``` {r}

library(ggplot2)
library(dplyr)


corpus <- read.csv("aisc.csv", header = TRUE)


corpus_avg <- corpus %>%
  group_by(arousal) %>%
  summarise(avg_danceability = mean(danceability, na.rm = TRUE))


ggplot(corpus, aes(x = arousal, y = danceability)) +
  geom_bin2d(bins = 10) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(
    title = "2D Histogram of Arousal vs Danceability",
    x = "Arousal (Energy Level)",
    y = "Danceability",
    fill = "Count"
  )



```

***

To get some more insights into the relationship between the two variables, I made another plot.

This plot visualizes the relationship between arousal (energy level) and danceability in AI-generated songs. Darker areas indicate where more songs exist, while lighter areas show less common combinations.

- No clear concentration: Songs are spread across various arousal and danceability levels, suggesting diverse musical characteristics.  
- High-density regions: Some areas (e.g., arousal ~4.5, danceability ~0.6) contain more songs, indicating common trends.  
- Rare combinations: Light yellow blocks highlight arousal-danceability pairs that occur less frequently.  

This suggests that the songs **do not follow a strict pattern**

